The Unicode Cookbook \linebreak for Linguists

### 2018-06-06 book/chapters/ipa_meets_unicode.tex
paperhivers up to page 25 or so...

Making things even more problematic, there is an old typographic tradition that
the double-story <a> uses a single-story <ɑ> in italics. This leads to the
unfortunate effect that even in many well-designed fonts the italics of <a> and
<ɑ> use the same glyph. For example, in Linux Libertine (the font of this book)
the italics of these characters are highly similar <\textit{a}> and
<\textit{ɑ}>, while in Charis SIL they are identical: <\textit{\charis{a}}> and
<\textit{\charis{ɑ}}>. If this distinction has to be kept upright in italics,
the only solution we can currently offer is to use \textsc{slanted} glyphs
(i.e.~artificially italicized glyphs) instead of real italics (i.e.~special
italics glyphs designed by a typographer).\footnote{For example, the widely used
IPA font Doulos SIL
(\url{http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL}) does not
have real italics. This leads some word-processing software, like Microsoft
Word, to produce slanted glyphs instead. That particular combination of font and
software application will thus lead to the desired effect distinguishing <a>
from <ɑ> in italics. However, note that when the text is transferred to another
font (i.e.~one that includes real italics) and/or to another software
application (like Apple Pages, which does not perform slanting), then this
visual appearance will be lost. In this case we are thus still in the
pre-Unicode situation in which the choice of font and rendering software
actually matters. The ideal solution from a linguistic point of view would be
the introduction of a new IPA code point for a different kind of <a>, which
explicitly specifies that it should still be rendered as a double-story
character when italicized. After informal discussion with various Unicode
players, our impression is that this highly restricted problem is not
sufficiently urgent to introduce even more <a> homographs in Unicode (which
already lead to much confusion, see Section~\ref{pitfall-homoglyphs}).} This 
approach was taken by the Language Science Press to distinguish 
between the two different orthographic <a>'s in Chakali in 
\cite{Brindle2017}.\footnote{\url{http://langsci-press.org/catalog/book/74}}

>>>>

Making things even more problematic, there is an old typographic tradition that
the double-story <a> uses a single-story <ɑ> in italics. This leads to the
unfortunate effect that even in many well-designed fonts the italics of <a> and
<ɑ> use the same glyph. For example, in Linux Libertine (the font of this book)
the italics of these characters are highly similar <\textit{a}> and
<\textit{ɑ}>, while in Charis SIL they are identical: <\textit{\charis{a}}> and
<\textit{\charis{ɑ}}>. If this distinction has to be kept upright in italics,
the only solution we can currently offer is to use \textsc{slanted} glyphs
(i.e.~artificially italicized glyphs) instead of real italics (i.e.~special
italics glyphs designed by a typographer).\footnote{For example, the widely used
IPA font Doulos SIL
(\url{https://software.sil.org/doulos/}) does not
have real italics. This leads some word-processing software, like Microsoft
Word, to produce slanted glyphs instead. That particular combination of font and
software application will thus lead to the desired effect distinguishing <a>
from <ɑ> in italics. However, note that when the text is transferred to another
font (i.e.~one that includes real italics) and/or to another software
application (like Apple Pages, which does not perform slanting), then this
visual appearance will be lost. In this case we are thus still in the
pre-Unicode situation in which the choice of font and rendering software
actually matters. The ideal solution from a linguistic point of view would be
the introduction of a new IPA code point for a different kind of <a>, which
explicitly specifies that it should still be rendered as a double-story
character when italicized. After informal discussion with various Unicode
players, our impression is that this highly restricted problem is not
sufficiently urgent to introduce even more <a> homographs in Unicode (which
already lead to much confusion, see Section~\ref{pitfall-homoglyphs}).} This 
approach was taken by the Language Science Press to distinguish 
between the two different orthographic <a>'s in Chakali in 
\cite{Brindle2017}.\footnote{\url{http://langsci-press.org/catalog/book/74}}

### 2018-06-06 book/chapters/pitfalls.tex
paperhivers up to page 25 or so...

A similar complaint to missing glyphs, discussed previously, is that while there
might be a glyph being displayed, it does not look right. There are two
reasons for unexpected visual display, namely automatic font substitution and
faulty rendering. Like missing glyphs, any such problems are independent from
the Unicode Standard. The Unicode Standard only includes very general
information about characters and leaves the specific visual display to others to
decide on. Any faulty display is thus not to be blamed on the Unicode
Consortium, but on a complex interplay of different mechanisms happening in a
computer to turn Unicode code points into visual symbols. We will only sketch a
few aspects of this complex interplay here.

Most modern software applications (like Microsoft Word) offer some approach to
\textsc{automatic font substitution}. This means that when a text is written in
a specific font (e.g.~Times New Roman) and an inserted Unicode character does not
have a glyph within this font, then the software application will automatically
search for another font to display the glyph. The result will be that this
specific glyph will look slightly different from the others. This mechanism
works differently depending on the software application, and mostly only limited
user influence is expected and little feedback is given, which might be rather
frustrating to font-aware users.\footnote{For example, Apple Pages does not give
any feedback that a font is being replaced, and the user does not seem to have
any influence on the choice of replacement (except by manually marking all
occurrences). In contrast, Microsoft Word does indicate the font replacement by
showing the name in the font menu of the font replacement. However, Word simply
changes the font completely, so any text written after the replacement is written in a
different font as before. Both behaviors leave much to be desired.}

The other problem with visual display is related to the so-called \textsc{font
rendering}. Font rendering refers to the process of the actual positioning of
Unicode characters on a page of written text. This positioning is actually a
highly complex problem, and many things can go wrong in the process. Well-known
rendering problems, like proportional glyph size or ligatures are reasonably
well understood. In contrast, the positioning of multiple diacritics relative to
a base character is still a widespread problem, even within the Latin script.
Especially when more than one diacritic is supposed to be placed above (or
below) each other, this often leads to unexpected effects in many modern
software applications. The problems arising in Arabic and in many southeast
Asian scripts (like Devanagari or Burmese) are even more complex.

To understand where any problems arise it is important to realize that there are
basically three different approaches to font rendering. The most widespread is
Adobe's and Microsoft's \textsc{OpenType} system. This approach makes it
relatively easy for font developers, as the font itself does not include all
details about the precise placement of individual characters. For those details,
additional script-descriptions are necessary. All of those systems can lead to
unexpected behavior.\footnote{For more details about OpenType, see
\url{http://www.adobe.com/products/type/opentype.html} and
\url{http://www.microsoft.com/typography/otspec/}. Additional systems for
complex text layout are, among others, Microsoft's DirectWrite
\url{https://msdn.microsoft.com/library/dd368038.aspx} and the open-source
project HarfBuzz \url{http://www.freedesktop.org/wiki/Software/HarfBuzz/}.}
Alternative systems are \textsc{Apple Advanced Typography} (AAT) and the
open-source \textsc{Graphite} system from the Summer Institute of Linguistics
(SIL).\footnote{More information about AAT can be found at
\url{https://developer.apple.com/fonts/}. \newline SIL's Graphite is described
in detail at
\url{http://scripts.sil.org/cms/scripts/page.php?site\_id=projects\&item\_id=graphite\_home}.}
In both of these systems, a larger burden is placed on the description inside
the font.

There is mostly no real solution to problems arising from faulty font rendering.
Switching to another software application that offers better handling is the
only real alternative, but this is normally not an option for daily work. The 
experience with rendering on the side of the software industry is developing 
quickly, so we can expect the situation to only get better. % In the meantime one 
% can try to correct faulty layout by tweaking baseline and/or kerning (when such 
% option are available).

@@

The Unicode code space is subdivided into blocks of contiguous code points. For
example, the block called \textsc{Cyrillic} runs from \uni{0400} till
\uni{04FF}. These blocks arose as an attempt at ordering the enormous amount of
characters in Unicode, but the ideas of blocks very quickly ran into problems.
First, the size of a block is fixed, so when a block is full, a new block will
have to be instantiated somewhere further in the code space. For example, this
led to the blocks \textsc{Cyrillic Supplement}, \textsc{Cyrillic Extended-A}
(both of which are also already full) and \textsc{Cyrillic Extended-B}. Second,
when a specific character already exists, then it is not duplicated in another
block, although the name of the block might indicate that a specific symbol
should be available there. In general, names of blocks are just an approximate
indication of the kind of characters that will be in the block.

@@

In general, finding a specific character in the Unicode Standard is often not
trivial. The names of the blocks can help, but they are not (and never were supposed
to be) a foolproof structure. It is not the goal nor aim of the Unicode
Consortium to provide a user interface to the Unicode Standard. If one often
encounters the problem of needing to find a suitable character, there are
various other useful services for end-users available.\footnote{The Unicode
website offers a basic interface to the code charts at
\url{http://www.unicode.org/charts/index.html}. As a more flexible interface, we
particularly like PopChar from Macility, available for both Macintosh and
Windows. There are also various free websites that offer search interfaces
to the Unicode code tables, like \url{http://unicode-search.net} or
\url{http://unicode-search.net}. A further useful approach for searching characters
using shape matching is \url{http://shapecatcher.com}.}

>>>>

A similar complaint to missing glyphs, discussed previously, is that while 
a glyph might be displayed, it does not look right. There are two
reasons for unexpected visual display, namely automatic font substitution and
faulty rendering. Like missing glyphs, any such problems are independent from
the Unicode Standard. The Unicode Standard only includes very general
information about characters and leaves the specific visual display to others to
decide on. Any faulty display is thus not to be blamed on the Unicode
Consortium, but on a complex interplay of different mechanisms happening in a
computer to turn Unicode code points into visual symbols. We will only sketch a
few aspects of this complex interplay here.

Most modern software applications (like Microsoft Word) offer some approach to
\textsc{automatic font substitution}. This means that when a text is written in
a specific font (e.g.~Times New Roman) and an inserted Unicode character does not
have a glyph within this font, then the software application will automatically
search for another font to display the glyph. The result will be that this
specific glyph will look slightly different from the others. This mechanism
works differently depending on the software application, only limited
user influence is usually expected and little feedback is given, which might be rather
frustrating to font-aware users.\footnote{For example, Apple Pages does not give
any feedback that a font is being replaced, and the user does not seem to have
any influence on the choice of replacement (except by manually marking all
occurrences). In contrast, Microsoft Word does indicate the font replacement by
showing the name in the font menu of the font replacement. However, Word simply
changes the font completely, so any text written after the replacement is written in a
different font as before. Both behaviors leave much to be desired.}

The other problem with visual display is related to the so-called \textsc{font
rendering}. Font rendering refers to the process of the actual positioning of
Unicode characters on a page of written text. This positioning is actually a
highly complex problem, and many things can go wrong in the process. Well-known
rendering difficulties, like proportional glyph size or ligatures, are reasonably
well understood. In contrast, the positioning of multiple diacritics relative to
a base character is still a widespread problem, even within the Latin script.
Especially when more than one diacritic is supposed to be placed above (or
below) each other, this often leads to unexpected effects in many modern
software applications. The problems arising in Arabic and in many southeast
Asian scripts (like Devanagari or Burmese) are even more complex.

To understand where any problems arise it is important to realize that there are
basically three different approaches to font rendering. The most widespread is
Adobe's and Microsoft's \textsc{OpenType} system. This approach makes it
relatively easy for font developers, as the font itself does not include all
details about the precise placement of individual characters. For those details,
additional script-descriptions are necessary. All such systems can lead to
unexpected behavior.\footnote{For more details about OpenType, see
\url{http://www.adobe.com/products/type/opentype.html} and
\url{http://www.microsoft.com/typography/otspec/}. Additional systems for
complex text layout are, among others, Microsoft's DirectWrite
\url{https://msdn.microsoft.com/library/dd368038.aspx} and the open-source
project HarfBuzz \url{http://www.freedesktop.org/wiki/Software/HarfBuzz/}.}
Alternative systems are \textsc{Apple Advanced Typography} (AAT) and the
open-source \textsc{Graphite} system produced and maintained by the Non-Roman Script Initiative of SIL International 
(SIL).\footnote{More information about AAT can be found at
\url{https://developer.apple.com/fonts/}. \newline Graphite is described
in detail at:
\url{http://scripts.sil.org/default}.}
In both of these systems, a larger burden is placed on the description inside
the font.

There is no real solution to problems arising from faulty font rendering.
Switching to another software application that offers better handling is the
only real alternative, but this is normally not an option for daily work. 
Font rendering is developing quickly in the software industry, so we can expect 
the situation to only get better. % In the meantime one 
% can try to correct faulty layout by tweaking baseline and/or kerning (when such 
% option are available).

@@

The Unicode code space is subdivided into blocks of contiguous code points. For
example, the block called \textsc{Cyrillic} runs from \uni{0400} till
\uni{04FF}. These blocks arose as an attempt at ordering the enormous number of
characters in Unicode, but the idea of blocks very quickly ran into problems.
First, the size of a block is fixed, so when a block is full, a new block will
have to be instantiated somewhere else in the code space. For example, this
led to the blocks \textsc{Cyrillic Supplement}, \textsc{Cyrillic Extended-A}
(both of which are already full) and \textsc{Cyrillic Extended-B}. Second,
when a specific character already exists, it is not duplicated in another
block, although the name of the block might indicate that a specific symbol
should be available there. In general, names of blocks are just an approximate
indication of the kind of characters that will be in the block.

@@

In general, finding a specific character in the Unicode Standard is often non-trivial. 
The names of the blocks can help, but they are not (and were never supposed
to be) a foolproof structure. It is neither the goal nor the aim of the Unicode
Consortium to provide a user interface to the Unicode Standard. If one often
encounters the problem of needing to find a suitable character, there are
various other useful services for end-users available.\footnote{The Unicode
website offers a basic interface to the code charts at
\url{http://www.unicode.org/charts/index.html}. As a more flexible interface, we
particularly like PopChar from Ergonis Software, available for both Macintosh and
Windows. There are also various free websites that offer search interfaces
to the Unicode code tables, like \url{http://unicode-search.net} or
\url{http://unicode-search.net}. A further useful approach for searching characters
using shape matching is \url{http://shapecatcher.com}.}
